Crowd density estimation from Wi-Fi positioning data in the Amsterdam Arena

### 2016-07-12 PaperOnSmoothing/draft-31052016.tex
abstract added

\begin{abstract}
\begin{comment}
We present the Amsterdam Arena project, which involves observing and managing crowd behaviour, using the Amsterdam Arena stadium as a living laboratory. The main scientific question we explore is how to detect anomalous behaviour in large crowds in real time. The main purpose is to be able to predict a possible crowd disaster and identify means to prevent it from happening. 
Human crowds are complex systems, and predicting or controlling their behaviour is challenging. Our approach involves three phases, firstly data collection from Wi-Fi and Bluetooth sensors in the stadium, secondly data analytics, and finally we aim to use simulation to make forecasts about the crowd dynamics. 
Here we present initial results on the data analytics and show how we can extract density maps from Wi-Fi positioning data. The technology we deploy is based on the Wi-Fi signals from smart phones. We use the existing network of Wi-Fi access points in the stadium, and capture probe signals from smart phones, which are processed and anonymised in alignment with privacy concerns. The positions of smart phones are reconstructed using received signal strengths and methods similar to trilateration. The data provides us with real-time information on spatial distributions of crowd density, which is an essential indicator of the criticality of crowd conditions. To visualise the spatiotemporal behaviour of crowd density in real-time we dynamically generate heat maps along a moving time interval. The heat maps are generated through the statistical modelling of the positioning data. The generation of dynamic heat maps allows us to detect and locate hot-spots of density where crowd conditions reach critical values that could possibly lead to a disaster.
\end{comment}
\end{abstract}

@@

***********************************From Sonja. nb: the text will be polished in the final phases, first I care only about putting all ideas on paper***************

@@

In our approach, which can complement the video-based analysis, we combat the three above mentioned issues by exploiting the ubiquity of smart phones, as in ((4) (5), (6) from proposal?). Our living laboratory is the Amsterdam Arena \cite{arena}. Our methodology consists of three steps.  First, we are using Wi-Fi routers in the stadium to gather the strengths of the wireless antena signals from individuals smart phones.  Second, we are estimating anonymously the positions of the visitors from the signal strengths. This technique has been presented in \cite{customers_Jan}, and is similar to trilateration, that is, the way the Global Positioning System (GPS) is implemented. However, due to signal interference and blockage in dense crowds, irregular internet packet rates, and systematic errors leading to multiple local optima, the positioning itself does not suffice to estimate the crowd density. Thus, in our third step, we aggregate the results from the positioning and apply statistical ensembling to obtain an estimation of the crowd density. With the first two steps, we are handling issues 1) and 2) mentioned above. With the third step, which is the main contribution of this paper, we are handling issue 3) -  that is, with our method the precision of estimation actually increases as the crowd density increases.

@@

************************************end from Sonja
\begin{comment}
\begin{itemize}

>>>>

\begin{abstract}

We present a new methodology for estimating crowd density during festivals. Our motivation comes from the fact that many crowd disasters have happened because of abnormal crowd density. We exploit the ubiquity of smart phones, without requiring participation from the crowd, and we use Amsterdam ArenA as a living laboratory. We build upon previous work, where the location of a visitor is estimated anonymously based on the strengths of the Wi-Fi signals received at the ArenA Wi-Fi routers,  by a method similar to trilateration,. However, because in dense crowds a Wi-Fi signal is partially absorbed and partially reflected by the human bodies, the localization based on signals strengths alone does not suffice to estimate the crowd density. Moreover, since we don't rely on the smart phones being connected to the Wi-Fi network, the signal rates are quite volatile, and the MAC addresses of the phones may change, depending on the vendor's privacy policy. In this paper, we address those issues. We borrow ideas from statistical mechanics to estimate the crowd density: rather than estimating the position of a visitor, we create a continuous probability distribution over all possible positions. We rely on the law of the large numbers, with adaptation to correlated variables, to estimate the crowd density by aggregating the individual distributions. The individual probability distribution turns gradually into a uniform distribution, with which we address the problem of outdated signals. Finally, we use the law of large numbers again and the "randomized address" tag to account for the fact that a portion of the MAC addresses are randomized. Thus, the advantage of our methodology when compared to similar methodologies is that our estimation becomes more (rather than less) precise when the number of people increases. This suits the purpose of being able to detect raised crowd density. 
We validate our methodology by comparing it to the ground truth obtained from security cameras. \marginpar{val. to be done}

\begin{comment}
We present the Amsterdam Arena project, which involves observing and managing crowd behaviour, using the Amsterdam Arena stadium as a living laboratory. The main scientific question we explore is how to detect anomalous behaviour in large crowds in real time. The main purpose is to be able to predict a possible crowd disaster and identify means to prevent it from happening. 
Human crowds are complex systems, and predicting or controlling their behaviour is challenging. Our approach involves three phases, firstly data collection from Wi-Fi and Bluetooth sensors in the stadium, secondly data analytics, and finally we aim to use simulation to make forecasts about the crowd dynamics. 
Here we present initial results on the data analytics and show how we can extract density maps from Wi-Fi positioning data. The technology we deploy is based on the Wi-Fi signals from smart phones. We use the existing network of Wi-Fi access points in the stadium, and capture probe signals from smart phones, which are processed and anonymised in alignment with privacy concerns. The positions of smart phones are reconstructed using received signal strengths and methods similar to trilateration. The data provides us with real-time information on spatial distributions of crowd density, which is an essential indicator of the criticality of crowd conditions. To visualise the spatiotemporal behaviour of crowd density in real-time we dynamically generate heat maps along a moving time interval. The heat maps are generated through the statistical modelling of the positioning data. The generation of dynamic heat maps allows us to detect and locate hot-spots of density where crowd conditions reach critical values that could possibly lead to a disaster.
\end{comment}
\end{abstract}

@@

\marginpar{begin from Sonja. nb: the text will be polished in the final phases, first I care only about putting all ideas on paper}

@@

In our approach, which can complement the video-based analysis, we combat the three above mentioned issues by exploiting the ubiquity of smart phones. Our living laboratory is the Amsterdam Arena \cite{arena}. Our methodology consists of three steps.  First, we are using Wi-Fi routers in the stadium to gather the strengths of the wireless antena signals from individuals smart phones.  Second, we are estimating anonymously the positions of the visitors from the signal strengths. This technique has been presented in \cite{customers_Jan}, and is similar to trilateration, that is, the way the Global Positioning System (GPS) is implemented. However, due to signal interference and blockage in dense crowds, irregular internet packet rates, and systematic errors leading to multiple local optima, the positioning itself does not suffice to estimate the crowd density. Thus, in our third step, we aggregate the results from the positioning and apply statistical ensembling to obtain an estimation of the crowd density. With the first two steps, we are handling issues 1) and 2) mentioned above. With the third step, which is the main contribution of this paper, we are handling issue 3) -  that is, with our method the precision of estimation actually increases as the crowd density increases.

@@

\marginpar{end from Sonja}
\begin{comment}
\begin{itemize}
